Surrogate waveform model using Gaussian process regression for an aligned-spin effective one body model of binary neutron star inspirals

### 2018-10-02 paper/paper.tex
Updates to PE section.

As a final test of the surrogate, we perform parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template. For the two aLIGO detectors we use the design sensitivity PSD~\cite{}, and for the advanced Virgo detectors we also use its design sensitivity PSD. We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$, corresponding to the MPA1 equation of state. The sky location and inclination angle are fixed at an arbitrary value, and the distance is chosen such that the optimal network signal-to-noise ratio (SNR) is either 30 or 60. We inject the waveforms into zero noise data so that the recovered parameters do not depend on the specific noise realization~\cite{}.

As discussed in more detail in~\cite{BNSPE}, the posterior $p(\theta | d)$ for the parameters $\theta$ is given by Bayes' theorem:
\begin{equation}
p(\theta | d) \propto p(\theta) \mathcal{L}(d | \theta),
\end{equation}
where $p(\theta)$ is the prior and $\mathcal{L}(d | \theta)$ is the likelihood. We choose for our priors the sky position and orientation uniform on the unit sphere. The distance is uniform in co-moving volume. The masses, spins, and tidal parameters are uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$. We sample this posterior with the parallel-tempered MCMC code in LALInference. We sample the data and waveform at \red{8192}~Hz such that the Nyquist frequency is 4096~Hz. \red{[[We note that published analyses for BNS systems typically use a sampling frequency of 4096~Hz. However, the \texttt{SEOBNRv4T} waveform model still has some signal above the corresponding Nyquist frequency of 2048~Hz (See Fig.~\ref{fig:hoff}).]]} We integrate the likelihood function in the interval $[20, 4096]$~Hz. Approximately \red{$10^x$} iterations are needed to produce \red{$\sim 5000$} statistically independent samples.

For data with zero noise, the fit between the injected waveform and the template is best when the likelihood is maximum. However, as a result of the priors that are bounded and not flat for some parameters, the posterior, and especially the marginalized posteriors, will not be maximized at the injected parameters even with a perfect template. We therefore use the difference between the parameters that maximize the likelihood $\theta_{\rm max}$ and the injected parameters $\theta_{\rm inj}$ as our estimate for the systematic error due to the surrogate.

Marginalized 1-dimensional posteriors for four key parameters are shown in Fig.~\ref{fig:pe}. These are the chirp mass $\mathcal{M} = (m_1 m_1)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$. The maximum value of the likelihood found for each run is dashed, and the injected value is solid. Because the sampled parameters do not necessarily contain the parameters for the maximum likelihood, the difference between these parameters $\Delta\theta_{\rm sys} = \theta_{\rm max} - \theta_{\rm inj}$ is an upper bound on the systematic error that results from using the surrogate as the template instead of the original \texttt{SEOBNRv4T} waveform. We can compare this systematic error to the statistical error from the finite sensitivity of the detectors. We will define this as half the distance between the 5\% and 95\% percentiles of each parameter: $\Delta\theta_{\rm stat} = (\theta_{95} - \theta_{5})/2$. Table~\ref{tab:syserrors} shows that the systematic error is small compared to the statistical errors even for the injection with SNR=60.

\begin{figure}[htb]
\centering
\includegraphics[width=0.39\textwidth]{chirpmass.pdf}\\
\includegraphics[width=0.39\textwidth]{q.pdf}\\
\includegraphics[width=0.39\textwidth]{chieff.pdf}\\
\includegraphics[width=0.39\textwidth]{tildelambda.pdf}
\caption{Marginalized PDFs for two waveform injections with zero noise. The parameters are identical except for the distances which are set such that the network SNRs are 30 (blue) or 60 (red). The shaded regions are bounded by the 5 and 95 percentiles. The vertical solid black line represents the injected parameter value. The vertical dashed lines represent the parameter value with maximum likelihood.}
\label{fig:pe}
\end{figure}

\begin{table}[htb]
\centering    
\caption{Estimated systematic errors due to surrogate model and statistical errors from finite detector sensitivity. The ratio $\Delta\theta_{\rm sys} / \Delta\theta_{\rm stat}$ is given for the two injections with network SNRs of 30 and 60.} 
\begin{tabular}{ccc}        
\hline
$\theta$ & SNR=30 & SNR=60 \\
\hline
$\mathcal{M}$ &   &  \\
$q$ &   &  \\
$\chi_{\rm eff}$ &   &  \\
$\tilde\Lambda$ &   &  \\
\hline
\end{tabular}
\label{tab:syserrors}
\end{table}

@@

BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the Institute for Nuclear Theory where many of the methods used here were discussed. The authors were supported by...

>>>>

As a final test of the surrogate accuracy, we perform parameter estimation runs where we inject the original, time-domain \texttt{SEOBNRv4T} waveform and recover the injected parameters with the \texttt{SEOBNRv4T\_surrogate} template. We use the design sensitivity PSDs for the two advanced LIGO detectors~\cite{Aasi:2013wya} and the advanced Virgo detector~\cite{virgoPSD}. We choose as our waveform parameters $m_1=m_2=1.4M_\odot$, $\chi_1=\chi_2=0.1$, and $\Lambda_1=\Lambda_2=1286$ corresponding to the equation of state named MS1b in~\cite{ReadLackey2009}. The sky location and inclination angle are fixed at an arbitrary value, and the distance is chosen such that the optimal network signal-to-noise ratio (SNR) is either 30 or 60. We inject the waveforms into zero noise data so that the recovered parameters do not depend on the specific noise realization~\cite{NissankeHolzHughes2010}.

As discussed in more detail in~\cite{BNSPE}, the posterior $p(\vec\vartheta | d)$ for the parameters $\vec\vartheta$ given the detector data $\vec d$ is given by Bayes' theorem:
\begin{equation}
p(\vec\vartheta | \vec d) \propto p(\vec\vartheta) \mathcal{L}(\vec d | \vec\vartheta),
\end{equation}
where $p(\vec\vartheta)$ is the prior and $\mathcal{L}(\vec d | \vec\vartheta)$ is the likelihood. We choose for our priors the sky position and orientation uniform on the unit sphere. The distance is uniform in co-moving volume. The masses, spins, and tidal parameters are uniform in the ranges $m_{1, 2} \in [0.7, 2]M_\odot$, $\chi_{1,2} \in [-0.5, 0.5]$ and $\Lambda_{1,2} \in [0, 5000]$. The mass ratio is restricted to the range $q\in[0.5, 1]$. We sample this posterior with the parallel-tempered MCMC code in \texttt{LALInference}~\cite{lal}. As was done in~\cite{BNSPE}, we sample the data and waveform at \red{4096}~Hz, such that the Nyquist frequency is \red{2048}~Hz, and integrate the likelihood function in the interval \red{$[20, 2048]$~Hz}. Approximately \red{$10^x$} iterations are needed to produce \red{$\sim 5000$} statistically independent samples. We note that the \texttt{SEOBNRv4T} waveform model still has some signal above \red{2048~Hz} (see Fig.~\ref{fig:hoff}). To test if this had any effect on the recovered posteriors, we also performed runs with a sampling frequency of \red{8192}~Hz and integrated the likelihood function in the interval \red{$[20, 4096]$}~Hz, but did not see a significant difference in marginalized posteriors.

Marginalized 1-dimensional posteriors for four key parameters are shown in Fig.~\ref{fig:pe}. These are the chirp mass $\mathcal{M} = (m_1 m_1)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$. We note that the peaks of the marginalized posteriors do not line up with the injected values. One reason is that the priors are not flat in these parameters. More importantly, because the posterior is bounded by the prior for many parameters, strong correlations cause a shift in other parameters. We therefore need another measure to examine the systematic error due to the surrogate.

For data with zero noise, the fit between the injected waveform and the template is best when the likelihood is maximum. We therefore use the difference $\Delta\vec\vartheta_{\rm sys} = \vec\vartheta_{\rm max} - \vec\vartheta_{\rm inj}$ between the parameters that maximize the likelihood $\vec\vartheta_{\rm max}$ and the injected parameters $\vec\vartheta_{\rm inj}$ as our estimate for the systematic error due to the surrogate. In Fig.~\ref{fig:pe}, the maximum sampled likelihood is dashed, and the injected value is solid. This difference is a conservative upper limit because the maximum sampled likelihood does not necessarily correspond to the actual maximum. 
%This is especially true for strongly correlated parameters such as the pair $q$ and $\chi_{\rm eff}$. 
Finally, we can compare this systematic error to the statistical error which we define as the 90\% credible interval shown by the shaded region. We find that the systematic error is significantly less that the statistical error, even for the SNR=60 signal.

%For data with zero noise, the fit between the injected waveform and the template is best when the likelihood is maximum. However, as a result of the priors that are bounded and not flat for some parameters, the posterior, and especially the marginalized posteriors, will not be maximized at the injected parameters even with a perfect template. We therefore use the difference between the parameters that maximize the likelihood $\vec\vartheta_{\rm max}$ and the injected parameters $\vec\vartheta_{\rm inj}$ as our estimate for the systematic error due to the surrogate. 
%
%Marginalized 1-dimensional posteriors for four key parameters are shown in Fig.~\ref{fig:pe}. These are the chirp mass $\mathcal{M} = (m_1 m_1)^{3/5}/(m_1+m_2)^{1/5}$, the mass ratio $q=m_2/m_1$, the effective spin $\chi_{\rm eff} = (m_1\chi_1+m_2\chi_2)/(m_1+m_2)$ and the effective tidal deformability parameter $\tilde\Lambda = (16/13) [(1+12q)\Lambda_1 + (12+q)q^4\Lambda_2]/(1+q)^5$. The maximum value of the likelihood found for each run is dashed, and the injected value is solid. Because the sampled parameters do not necessarily contain the parameters for the maximum likelihood, the difference between these parameters $\Delta\vec\vartheta_{\rm sys} = \vec\vartheta_{\rm max} - \vec\vartheta_{\rm inj}$ is an upper bound on the systematic error that results from using the surrogate as the template instead of the original \texttt{SEOBNRv4T} waveform. We can compare this systematic error to the statistical error $\Delta\vec\vartheta_{\rm stat}$ from the finite sensitivity of the detectors. Here, we will $\Delta\vec\vartheta_{\rm stat}$ as half the width of the 90\% credible interval. Table~\ref{tab:syserrors} shows that the systematic error is small compared to the statistical errors even for the injection with SNR=60.

\begin{figure}[htb]
\centering
\includegraphics[width=0.39\textwidth]{chirpmass.pdf}\\
\includegraphics[width=0.39\textwidth]{q.pdf}\\
\includegraphics[width=0.39\textwidth]{chieff.pdf}\\
\includegraphics[width=0.39\textwidth]{tildelambda.pdf}
\caption{Marginalized PDFs for two waveform injections with zero noise. The parameters are identical except for the distances which are set such that the network SNRs are 30 (blue) or 60 (red). The shaded regions give the 90\% credible interval. The vertical solid black line represents the injected parameter value. The vertical dashed lines represent the parameter value with maximum likelihood.}
\label{fig:pe}
\end{figure}

%\begin{table}[htb]
%\centering    
%\caption{Estimated systematic errors due to surrogate model and statistical errors from finite detector sensitivity. The ratio $\Delta\vec\vartheta_{\rm sys} /\Delta\vec\vartheta_{\rm stat}$ is given for the two injections with network SNRs of 30 and 60.} 
%\begin{tabular}{ccc}        
%\hline
%$\vartheta$ & SNR=30 & SNR=60 \\
%\hline
%$\mathcal{M}$ &  $<$ & $<$ \\
%$q$ & $<$  & $<$ \\
%$\chi_{\rm eff}$ & $<$  & $<$ \\
%$\tilde\Lambda$ &  $<$ & $<$ \\
%\hline
%\end{tabular}
%\label{tab:syserrors}
%\end{table}

@@

BL thanks the participants of the Bayesian Methods in Nuclear Physics workshop at the University of Washington for useful discussions at the start of this work. The authors were supported by... \red{[[Get AEI funding info.]]}
